---
title: "Parametric models with {icpack}"
author: "Hein Putter & Paul Eilers"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parametric models with {icpack}}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding[UTF-8]{inputenc}
---


```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width = 6, fig.height = 4,
                      echo = TRUE, warning = FALSE, message = FALSE)
```

# Introduction

The purpose of this vignette is to show how {`icpack`} can be used to fit certain parametric models, namely with constant hazards, Gompertz models and Weibull models. We will use the ovarian cancer data, because with right-censored data we can easily check the results of {`icpack`} with other methods. Those other methods do not extend to interval-censored data, while {`icpack`} does.

For information on the ovarian cancer data we refer to the vignette "Smooth modeling of right-censored survival data with {`icpack`}", or the help file. First the package needs to be loaded.

```{r icpack}
library(icpack)
library(ggplot2)
```

# Constant hazards

Using `icfit` it is possible to fit proportional hazards models to right-censored and interval-censored data with constant baseline hazards. It is of interest to start with a model without covariates and right-censored data, since then we know what the maximum likelihood of the underlying constant hazard is, and we also have a straightforward expression for its confidence interval. Here is how the model is fit using `icfit`.

```{r icfit_cst}
ic_fit_cst <- icfit(Surv(time, d) ~ 1, data = Ova,
                    bdeg=1, pord=1, lambda=1e+06, update_lambda=FALSE)
ic_fit_cst
```

We use `bdeg=1` to use B-splines of degree 1 (constants), `pord=1` to penalize first order differences of coefficients of the B-splines, `lambda=1e+06` to use an extremely high penalty, and `update_lambda=FALSE` to make sure that lambda is not updated in the process. Note that the effective dimension of the log-hazard is effectively 1, so the log-hazard and hazard are constant. A plot of the hazard shows a straight line.

```{r plot_icfit_cst}
plot(ic_fit_cst)
```

The value fo the hazard and its 95% confidence interval can be seen by printing the result of `predict`.

```{r predict_icfit_cst}
pic_fit_cst <- predict(ic_fit_cst)[, c("time", "haz", "sehaz", "lowerhaz", "upperhaz")]
head(pic_fit_cst)
tail(pic_fit_cst)
```

We see that the hazard is estimated as $\hat{\lambda} = 0.000747$, with 95\% confidence interval about 0.000662 -- 0.000842. The estimate and confidence interval are obtained using the general theory outlined in Putter, Gampe \& Eilers (2023). Nevertheless, they coincide closely with standard theory, which says that $\hat{\lambda}_{\textrm{MLE}} = D/T$, where $D = \sum_{i=1}^n d_i$ is the total number of events and $T = \sum_{i=1}^n t_i$ is the total follow-up time in the data. (Here $(t_i, d_i)$ for subject $i=1,\ldots,n$ are the usual time and status indicators in right-censored time-to-event data.) Moreover, the variance of the log of $\hat{\lambda}_{\textrm{MLE}}$ is given by $1/D$, so a 95\% confidence interval for $\lambda$ is given by $\exp( \log(\hat{\lambda}_{\textrm{MLE}}) \pm 1.96/\sqrt{D} )$. In our data this would yield

```{r DTrate}
D <- sum(Ova$d)
T <- sum(Ova$time)
rate <- D / T
lograte <- log(rate)
selograte <- 1 / sqrt(D)
lowerlograte <- lograte - qnorm(0.975) * selograte
upperlograte <- lograte + qnorm(0.975) * selograte
data.frame(D=D, T=T, rate=rate, lower=exp(lowerlograte), upper=exp(upperlograte))
```

We see that this is quite close to what `icfit` has given us.

We can go one step further with `icfit` and add the covariate effects to the constant baseline hazard.

```{r icfit_cst_covs}
ic_fit_cst <- icfit(Surv(time, d) ~ Diameter + FIGO + Karnofsky, data = Ova,
                    bdeg=1, pord=1, lambda=1e+06, update_lambda=FALSE)
ic_fit_cst
```

Also this model can be fitted in an alternative way, namely using direct Poisson regression. To achieve this `d` is considered the outcome variable and the logarithm of the follow-up time is added as an offset. Here is the code to run it.

```{r Poissonreg}
Ova$logtime <- log(Ova$time)
pois_fit <- glm(d ~ Diameter + FIGO + Karnofsky + offset(logtime), family=poisson(), data = Ova)
spois_fit <- summary(pois_fit)
spois_fit
```

We see that the estimates of the covariate effects are quite similar to the estimates from `ic_fit_cst`. Moreover the `(Intercept)` term is the logarithm of the (constant) baseline hazard. If we exponeniate it the corresponding baseline hazard and 95\% confidence interval are given by

```{r baselineCI}
lograte <- spois_fit$coefficients[1, 1]
selograte <- spois_fit$coefficients[1, 2]
lowerlograte <- lograte - qnorm(0.975) * selograte
upperlograte <- lograte + qnorm(0.975) * selograte
data.frame(rate=exp(lograte), lower=exp(lowerlograte), upper=exp(upperlograte))
```

The result from `icfit` is visualised as

```{r plot_icfit_cst_covs}
plot(ic_fit_cst, xlab="Days since randomisation", ylab="Hazard",
     title="Constant baseline hazard") + ylim(0, 0.0052)
```

Naturally, direct Poisson regression using `glm` is preferable to `icfit` in case of constant hazards both in terms of speed and accuracy, but it is comforting to see  that `icfit` does give rapid and reliable results also here. Moreover, Poisson regression cannot be used directly with interval-censored data.

Precision in `icfit` can be increased (at the cost of extra computation time) by increasing `nt`. If we remove the covariates again, we set `nt` to 500, and look at the hazard and its 95% confidence interval, which again can be seen by printing the result of `predict`, we do see that the results are quite a bit closer to that of the occurrence over exposure theoretical ones.

```{r icfit_nt500}
ic_fit_cst <- icfit(Surv(time, d) ~ 1, data = Ova, nt=500,
                    bdeg=1, pord=1, lambda=1e+06, update_lambda=FALSE)
pic_fit_cst <- predict(ic_fit_cst)[, c("time", "haz", "sehaz", "lowerhaz", "upperhaz")]
head(pic_fit_cst)
tail(pic_fit_cst)
```

Note that only with right-censored data we can use the occurrence over exposure theory and Poisson GLM. Importantly, {`icpack`} can also be used to fit models (with and without covariates) with constant (baseline) hazards for interval-censored data, while there is no standard alternative for this situation.

# "Gompertz" models

When we set the degree of the B-splines to 1 and the order of the penalty to 2, and take a very large value of lambda, we get a model for which the log-hazard is a straight line. The corresponding hazard is of the form $h(t) = a e^{bt}$. This corresponds to the hazard of a Gompertz model, provided $b>0$. Gompertz rates are often used to represent rates of mortality; when $b<0$ then this implies that very large, or indeed infinite values are not uncommon, since then $\int_0^\infty h(t) dt$ is finite. For modeling human mortality this obviously is a problem, but for many other practical purposes we may ignore this issue. But implicitly allowing for negative $b$ by `icfit` is the reason for referring to "Gompertz" rather than Gompertz models.

Here is the fit of the "Gompertz" model using `icfit`.

```{r icfit_gompertz}
ic_fit_gompertz <- icfit(Surv(time, d) ~ 1, data = Ova,
                         bdeg=1, pord=2, lambda=1e+06, update_lambda=FALSE)
ic_fit_gompertz
```

We see that the effective dimension of the log-hazard is 2, so a straight
line. Below is a plot of the hazard.

```{r plot_gompertz}
plot(ic_fit_gompertz) + ylim(0, 0.00125)
```

The fact that the log-hazard is a straight line is easier to see after transforming the y-axis.

```{r plot_gompertz_logy}
plot(ic_fit_gompertz) + scale_y_continuous(trans='log10')
```

Note that the "Gompertz" hazard rate is implicitly obtained by setting the order of the penalty to 2. From the result of `icfit` the values of $a$ and $b$ and their standard errors can only be derived indirectly.

## Weibull models

Also Weibull models can be fitted with `icfit` in the same indirect way, by setting `bdeg`, `pord`, and `lambda`, with one additional trick, namely time warping. The hazard of a Weibull distribution is of the form $h(t) = \alpha \lambda t^{\alpha-1}$, with $\lambda > 0$. If we take `pord=2`, `lambda = 1e+06`, and work with $\log(t)$ instead of $t$ itself, the form of the log-hazard would be $\beta_0 + \beta_1 \log(t)$, which corresponds to a Weibull hazard with $\alpha = \beta_1 + 1$, $\lambda = e^{\beta_0} / (\beta_1 + 1)$.

Here is the fit of the Weibull model with `icfit`. This works fine with our present way of using time in days; `time=1` corresponds to 1 day after randomisation. When time is in years, the logarithm of time would be negative for all events and censorings before one year. In that case adding a constant, or alternatively, defining `logtime` below as `log(Ova$time * 365.25)` would solve the issue.

```{r icfit_weibull}
Ova$logtime <- log(Ova$time)
ic_fit_weibull <- icfit(Surv(logtime, d) ~ 1, data = Ova,
                         bdeg=1, pord=2, lambda=1e+06, update_lambda=FALSE)
ic_fit_weibull
```

Here is a plot of the hazard with log-transformed y-axis. Time on the x-axis actually refers to the logarithm (base 10) of the original time, so 3 on the x-axis corresponds to $t=1000$.

```{r plot_icfit_weibull}
plot(ic_fit_weibull) + scale_y_continuous(trans='log')
```

To see everything on the original time scale and to facilitate future comparison with a direct parametric fit using `survreg`, let us extract the `data` elements of the plot, and explicitly transform time back into the original time.

We need to be careful though. Note that the estimated hazard with transformed time measures the expected number of events per *transformed* time unit, not in the original time unit. So we need not only back transform time, but also the hazard itself. If $\tilde{h}(u)$ denotes the hazard in transformed time $u = g(t)$, then in the original time scale $t = f(u) = g^{-1}(u)$ we have that $h(t) = \tilde{h}(u) / f^\prime(u)$. With $u = g(t) = \log(t)$, $t = f(u) = \exp(u)$, we get $h(t) = \tilde{h}(u) / \exp(u) = \tilde{h}(\log(t)) / t$.

```{r pweib}
pw <- plot(ic_fit_weibull)$data
head(pw)[, 1:5]
tail(pw)[, 1:5]
pw$u <- pw$time
pw$time <- exp(pw$u)
pw$haz <- pw$haz / pw$time
pw$lowerhaz <- pw$lowerhaz / pw$time
pw$upperhaz <- pw$upperhaz / pw$time
head(pw)[, 1:5]
tail(pw)[, 1:5]
```

Here is a plot of the result.

```{r plot_pweib}
plt <- ggplot(pw) +
  geom_ribbon(aes(x = pw$time, ymin = pw$lowerhaz, ymax = pw$upperhaz),
              fill = 'lightgrey') +
  geom_line(aes(x = pw$time, y = pw$lowerhaz), colour = 'blue') +
  geom_line(aes(x = pw$time, y = pw$upperhaz), colour = 'blue') +
  geom_line(aes(x = pw$time, y = pw$haz), size = 1, colour = 'blue', lty = 1) +
  ylim(0, 0.002) + 
  ggtitle("Weibull hazard fit") + xlab("Days since randomisation") +
  ylab("Hazard") + theme_bw() + theme(plot.title = element_text(hjust = 0.5))
plt
```

Let us compare this now with `survreg`. It is most convenient to use the wrapper provided in `flexsurvreg`, because it uses a parametrisation close to ours. The parametrisation in `flexsurvreg` is of the form $S(t) = \exp(-(t/\mu^\prime)^{\alpha^\prime})$, compared to our $S(t) = \exp(-\lambda t^{\alpha})$, from which we see that $\alpha = \alpha^\prime$ and $\lambda = 1 / (\mu^\prime)^{\alpha^\prime}$. We add the fit from `survreg` with a dashed line to the original plot.

```{r flexsurvreg}
library(flexsurv)
fsr <- flexsurvreg(Surv(time, d) ~ 1, data = Ova, dist = "weibull")
fsr

alp <- exp(fsr$coef[1])
mu <- exp(fsr$coef[2])
lam <- 1 / mu^alp
hw <- function(t, alp, lam) alp * lam * t^(alp - 1)
maxt <- max(Ova$time)
pw$haz_survreg <- hw(pw$time, alp, lam)

plt <- plt +
  geom_line(aes(x = pw$time, y = pw$haz_survreg), size = 1, lty = 2)
plt
```

The fit is quite similar, although not entirely equivalent.

## References

Putter, H., Gampe, J., Eilers, P. H. (2023). Smooth hazard regression for interval censored data. _Submitted for publication_.